Final project

### 2025-03-14 products/manuscript/Manuscript.qmd
created a different manuscript for the unfinished project codes
Create a new qmd file in the manuscript folder within the products folder to transfer all the codes for the unfinished final projects.

@@

print(cardio_df)

view(cardio_df)

>>>>

refernece: 
1- https://www.heart.org/en/health-topics/high-blood-pressure/understanding-blood-pressure-readings
1- https://www.mayoclinic.org/diseases-conditions/prehypertension/symptoms-causes/syc-20376703
2- https://my.clevelandclinic.org/health/articles/11920-cholesterol-numbers-what-do-they-mean
2- https://www.hopkinsmedicine.org/health/treatment-tests-and-therapies/lipid-panel
3- https://pmc.ncbi.nlm.nih.gov/articles/PMC6376358/
3- https://www.ahajournals.org/doi/pdf/10.1161/01.STR.25.9.1820

@@

# Transform variables into suitable types and label them according to the data dictionary to enhance readability.
cardio_df <- cardio_df %>%
  mutate(
    gender = factor(gender, levels = c(0, 1), labels = c("Female", "Male")), 
    chestpain = factor(chestpain, levels = c(0, 1, 2, 3),
                       labels = c("Typical Angina", "Atypical Angina", "Non-anginal Pain", "Asymptomatic")),
    fastingbloodsugar = factor(fastingbloodsugar, levels = c(0, 1), labels = c("False", "True")),
    restingrelectro = factor(restingrelectro, levels = c(0, 1, 2),
                             labels = c("Normal", "ST-T Abnormality", "Left Ventricular Hypertrophy")),
    exerciseangia = factor(exerciseangia, levels = c(0, 1), labels = c("No", "Yes")),
    slope = factor(slope, levels = c(1, 2, 3), labels = c("Upsloping", "Flat", "Downsloping")),
    target = factor(target, levels = c(0, 1), labels = c("No Heart Disease", "Heart Disease"))
  )

# Part 2. Exploratory Data Analysis (EDA):
```{r}
# Calculate summary statistics for selected variables

summary_stats <- cardio_df %>% 
  summarise(
    # Age
    Age_min = min(age, na.rm = TRUE),
    Age_max = max(age, na.rm = TRUE),
    Age_mean = mean(age, na.rm = TRUE),
    Age_sd = sd(age, na.rm = TRUE),

    # Resting Blood Pressure
    RestingBP_min = min(restingBP, na.rm = TRUE),
    RestingBP_max = max(restingBP, na.rm = TRUE),
    RestingBP_mean = mean(restingBP, na.rm = TRUE),
    RestingBP_sd = sd(restingBP, na.rm = TRUE),

    # Serum Cholesterol
    SerumCholesterol_min = min(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),
    SerumCholestrol_max = max(serumcholestrol, na.rm = TRUE),
    SerumCholestrol_mean = mean(serumcholestrol, na.rm = TRUE),
    SerumCholestrol_sd = sd(serumcholestrol, na.rm = TRUE),

    # Maximum Heart Rate
    MaxHeartRate_min = min(maxheartrate, na.rm = TRUE),
    MaxHeartRate_max = max(maxheartrate, na.rm = TRUE),
    MaxHeartRate_mean = mean(maxheartrate, na.rm = TRUE),
    MaxHeartRate_sd = sd(maxheartrate, na.rm = TRUE),

    # Oldpeak
    Oldpeak_min = min(oldpeak[oldpeak > 0], na.rm = TRUE),
    Oldpeak_max = max(oldpeak, na.rm = TRUE),
    Oldpeak_mean = mean(oldpeak, na.rm = TRUE),
    Oldpeak_sd = sd(oldpeak, na.rm = TRUE),
  )

# Print the summary statistics
print(summary_stats)

# Histogram and density overlay of age distribution
ggplot(cardio_df, aes(x = age)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "steelblue", alpha = 0.5) +
  geom_density(color = "darkblue", linewidth = 1) +
  labs(title = "Age Distribution", x = "Age", y = "Density") +
  scale_x_continuous(breaks = seq(0, max(cardio_df$age, na.rm = TRUE) + 10, by = 10)) +
  theme_bw()

# Histogram and density overlay of serum cholestrol (excluding zeros)
ggplot(subset(cardio_df, serumcholestrol > 0), aes(x = serumcholestrol)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "forestgreen", alpha = 0.5) +
  geom_density(color = "darkgreen", linewidth = 1) +
  labs(title = "Serum Cholesterol Distribution", 
       x = "Serum Cholesterol (mg/dL)", y = "Density") +
  theme_minimal()

# Histogram and density overlay of blood pressure (excluding zeros)
ggplot(subset(cardio_df, restingBP > 0), aes(x = restingBP)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "maroon", alpha = 0.5) +
  geom_density(color = "darkred", linewidth = 1) +
  labs(title = "Resting Blood Pressure Distribution", 
       x = "Resting Blood Pressure (mmHG)", y = "Density") +
  theme_bw()

# Part 3, Statistical analysis: 
```{r}

# Filter data for younger individuals (age < 40) with normal cholesterol levels, elevated BP and elevated oldpeak
cleaned_df <- cardio_df %>%
  filter(age < 40, serumcholestrol > 0,serumcholestrol < 200) %>%  # Normal if serumcholestrol < 200 mg/dL
  mutate(
    elevated_BP = case_when(
      restingBP < 120 ~ "Low BP",
      between(restingBP, 120, 129) ~ "Elevated",  # Elevated blood pressure: restingBP between 120 and 129
      restingBP > 129 ~ "High BP"
    ),
    elevated_oldpeak = if_else(oldpeak > 1, "Elevated", "Normal")  # Elevated ST depression: oldpeak > 1
  ) %>% 
  select(age, chestpain, restingBP, serumcholestrol, fastingbloodsugar, maxheartrate, oldpeak, target, elevated_BP, elevated_oldpeak)  # Select variables of interest

# Summarize the counts by blood pressure status, ST depression status, and CVD risk
summary_table <- cleaned_df %>%
  group_by(elevated_BP, elevated_oldpeak, target) %>%
  summarise(count = n(), .groups = "drop")

print(summary_table)

# Visualize CVD risk by blood pressure and ST depression status
ggplot(cleaned_df, aes(x = elevated_BP, fill = target)) +
  geom_bar(position = "dodge") +
  facet_wrap(~ elevated_oldpeak) +
  labs(title = "CVD Risk by Blood Pressure and ST Depression Status\n(Individuals Age < 40 with Normal Cholesterol)",
       x = "Blood Pressure Status",
       y = "Count",
       fill = "Heart Disease") +
  theme_minimal()

### 2025-03-14 products/manuscript/Unfinished project codes.qmd
created a different manuscript for the unfinished project codes
Create a new qmd file in the manuscript folder within the products folder to transfer all the codes for the unfinished final projects.

---
title: "Yaqubi-MADA-projet"
author: ""
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
---

The structure below is one possible setup for a manuscript, or a general data analysis project (including the course project). Adjust as needed. You don't need to have exactly these sections, but the content covering those sections should be addressed.

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information. You can switch to other formats, like html or pdf. See [the Quarto documentation](https://quarto.org/) for other formats.

@@

**Authors**

-   First Autor$^{1,2,*}$ (ORCID: 0000-0000-1234-5678)
-   Second Author$^{3, *}$
-   Third Autor$^{3, \dagger}$
-   Last Author$^{1, 2, \land}$

**Author affiliations**

1.  College of Public Health, University of Georgia, Athens, GA, USA.
2.  Another Affiliation, Athens, GA, USA.
3.  Yet another Affiliation.

$*$ These authors contributed equally to this work.

$\land$ Corresponding author: some\@email.com

$\dagger$ Disclaimer: The opinions expressed in this article are the author's own and don't reflect their employer.

{{< pagebreak >}}

# Summary/Abstract

*Write a summary of your project.*

{{< pagebreak >}}

# Introduction

@@

Coronary artery disease (CAD), also known as coronary heart disease or ischemic heart disease. It’s also what most people mean when they use the general term “heart disease.” It is a common and serious condition where plaque buildup in the coronary arteries restricts blood flow to the heart. This gradual narrowing, caused by atherosclerosis, often develops over years without symptoms until a heart attack occurs, making CAD a "silent killer." Symptoms, when present, include chest pain and shortness of breath. As the leading cause of death in the U.S. and globally, CAD claimed 375,500 lives in the U.S. in 2021 and affects over 18 million American adults.

@@

Do younger individuals (e.g., age \< 40) with normal cholesterol levels (within "healthy" range) still face CVD risk if they have elevated blood pressure or ST depression (oldpeak)?

@@

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].

{{< pagebreak >}}

# Methods

@@

*Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement.*

## Schematic of workflow

Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it). @fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender. We store those figures in the `assets` folder.

```{r}
#| label: fig-schematic
#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
#| echo: FALSE
knitr::include_graphics(here("assets","antigen-recognition.png"))
```

@@

## Data import and cleaning

@@

head(cardio_df)

colnames(cardio_df)

dim(cardio_df)

summary(cardio_df)

glimpse(cardio_df)

gg_miss_var(cardio_df)

@@

# Part 2. Exploratory Data Analysis (EDA):
```{r}
# Calculate summary statistics for selected variables

@@

# Part 3, Statistical analysis: 
```{r}

# Filter data for younger individuals (age < 40) with normal cholesterol levels, elevated BP and elevated oldpeak
cleaned_df <- cardio_df %>%
  filter(age < 40, serumcholestrol > 0,serumcholestrol < 200) %>%  # Normal if serumcholestrol < 200 mg/dL
  mutate(
    elevated_BP = case_when(
      restingBP < 120 ~ "Low BP",
      between(restingBP, 120, 129) ~ "Elevated",  # Elevated blood pressure: restingBP between 120 and 129
      restingBP > 129 ~ "High BP"
    ),
    elevated_oldpeak = if_else(oldpeak > 1, "Elevated", "Normal")  # Elevated ST depression: oldpeak > 1
  ) %>% 
  select(age, chestpain, restingBP, serumcholestrol, fastingbloodsugar, maxheartrate, oldpeak, target, elevated_BP, elevated_oldpeak)  # Select variables of interest

@@

# Visualize CVD risk by blood pressure and ST depression status
ggplot(cleaned_df, aes(x = elevated_BP, fill = target)) +
  geom_bar(position = "dodge") +
  facet_wrap(~ elevated_oldpeak) +
  labs(title = "CVD Risk by Blood Pressure and ST Depression Status\n(Individuals Age < 40 with Normal Cholesterol)",
       x = "Blood Pressure Status",
       y = "Count",
       fill = "Heart Disease") +
  theme_minimal()

@@

*Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along.*

## Statistical analysis

*Explain anything related to your statistical analyses.*

{{< pagebreak >}}

# Results

## Exploratory/Descriptive analysis

*Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project.*

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below. I generally recommend the `here` package.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS("../../results/tables/summarytable.rds")
knitr::kable(resulttable)
```

## Basic statistical analysis

*To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by gender."
#| echo: FALSE
knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
```

## Full analysis

*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here.*

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
knitr::kable(resulttable2)
```

{{< pagebreak >}}

# Discussion

## Summary and Interpretation

*Summarize what you did, what you found and what it means.*

## Strengths and Limitations

*Discuss what you perceive as strengths and limitations of your analysis.*

## Conclusions

*What are the main take-home messages?*

*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like.

{{< pagebreak >}}

# References

>>>>

---
title: "Final project"
author: "Murtaza Yaqubi"
format: html
editor: visual
---

@@

@@

Coronary artery disease (CAD), also known as coronary heart disease or ischemic heart disease. It’s also what most people mean when they use the general term “heart disease”. It is a common and serious condition where plaque buildup in the coronary arteries restricts blood flow to the heart. This gradual narrowing, caused by atherosclerosis, often develops over years without symptoms until a heart attack occurs, making CAD a "silent killer." Symptoms, when present, include chest pain and shortness of breath. As the leading cause of death in the U.S. and globally, CAD claimed 375,500 lives in the U.S. in 2021 and affects over 18 million American adults.

@@

Do individuals younger than 50 with normal cholesterol levels (within "healthy" range) still face CVD risk if they have elevated blood pressure or ST depression (oldpeak)?

@@

## Methods

@@

@@

## Part 1, Data import and cleaning

@@

head(cardio_df)  # View the column names

colnames(cardio_df)  # View the column names

dim(cardio_df)  # View the dimensions of the data

summary(cardio_df)  # View the summary of the data

glimpse(cardio_df)  # Get a glimpse of the dat

gg_miss_var(cardio_df)  # Ilustrate the number of NA's for each of the variables

@@

print(cardio_df)
```

## Part 2. Exploratory Data Analysis (EDA):

```{r}
# Calculate summary statistics for selected variables

@@

## Part 3, Statistical analysis:

```{r}

# Filter data for younger individuals (age < 40) with normal cholesterol levels, elevated BP and elevated oldpeak
cleaned_df <- cardio_df %>%
  filter(age < 50, serumcholestrol > 0,serumcholestrol < 200) %>%  # Normal if serumcholestrol < 200 mg/dL
  mutate(
    elevated_BP = if_else(between(restingBP, 120, 129), "Elevated", "Otherwise"),  # Elevated blood pressure: restingBP between 120 and 129
    elevated_oldpeak = if_else(oldpeak > 1, "Elevated", "Normal")  # Elevated ST depression: oldpeak > 1
  ) %>% 
  select(age, chestpain, restingBP, serumcholestrol, fastingbloodsugar, maxheartrate, oldpeak, target, elevated_BP, elevated_oldpeak)   # Select variables of interest

@@

# Visualize CVD risk by blood pressure and ST depression status
ggplot(cleaned_df, aes(x = elevated_BP, fill = target)) +
  geom_bar(position = "dodge") +
  facet_wrap(~ elevated_oldpeak) +
  labs(title = "CVD Risk by Blood Pressure and ST Depression Status\n(Individuals Age < 50 with Normal Cholesterol)",
       x = "Blood Pressure Status",
       y = "Count",
       fill = "Heart Disease") +
  theme_minimal()

@@
